---
title: "spitzemap_construction"
author: "Chris Mathy"
email: "chris.mathy@ucsf.edu"
date: "3/22/2019"
output: html_document
---

import packages
```{r echo=T, results='hide'}
library(tidyverse)
```

Read in the Gsp1 data, output from MATLAB
- The data is processed to have six labels for each strain (query or array) to match the labels
  used for the SGA preprocessing.
```{r}
name_to_ORF <- list()
name_to_ORF$array_allele_name <- read_delim('gsp1_pEMAP_avg_merged_gene_names.txt', delim='\t') %>% colnames()
name_to_ORF$array_ORF <- read_delim('gsp1_pEMAP_avg_merged.txt', delim='\t') %>% colnames()
name_to_ORF <- as.data.frame(name_to_ORF)

gsp1_pEMAP <-
  read_delim('gsp1_pEMAP_avg_merged_gene_names.txt', delim='\t') %>% 
  rename(query_allele_name = Gene) %>%
  separate(query_allele_name, into = c('query_name', 'query_mutant'),
           sep = ' - ', extra = "drop", remove = FALSE) %>% 
  mutate(query_ORF = 'YLR293C', query_descriptor = NA) %>% 
  unite('query_strain_id', c('query_ORF','query_mutant'), remove = FALSE) %>% 
  gather(-starts_with('query'), key = 'array_allele_name', value = 'score', na.rm = TRUE) %>%
  left_join(name_to_ORF, by = 'array_allele_name') %>% 
  separate(array_allele_name, into = c('array_name', 'array_descriptor'),
           sep = ' - ', extra = "drop", remove = FALSE) %>% 
  mutate('array_mutant' = NA,
         'array_strain_id' = paste(array_ORF, 'emap', sep = '_'),
         'interaction_network' = 'gsp1_pEMAP',
         'pvalue' = NA) %>%
  select(starts_with('query'), starts_with('array'), score, pvalue, interaction_network)
```

import SGA data and prepare it for concatenation
```{r}
load('Data/SGA_scaling/SGA_2016_full_scaled.rda')

SGA_to_combine <-
  SGA_scaled_to_EMAP %>% 
  mutate(interaction_network = paste(interaction_network, 'SGA', sep = '_')) %>% 
  select(interaction_network, starts_with('query'), starts_with('array'),
         -ends_with('smf'), -arraytype_temp, score_scaled_to_emap, pvalue) %>% 
  rename('score' = score_scaled_to_emap)
```

Combine pEMAP with SGA to make spitzemap, then save
```{r}
spitzemap <- bind_rows(gsp1_pEMAP, SGA_to_combine)
save(spitzemap, file = 'spitzemap.rda')
```

Prepare the "spitzemapko", which only includes KOs as arrays, for use in
computing correlatons to Gsp1 mutant profiles.
```{r}
spitzemapko <-
  spitzemap %>% 
  filter(interaction_network %in% c('gsp1_pEMAP','ExN_SGA','NxN_SGA'),
         !grepl('supp', query_allele_name), # exclude strains with suppressive mutations
         !grepl('supp', array_allele_name),
         !grepl('_tsa', array_strain_id)) # exclude ts array genes (not in gsp1 emap library anyway)

save(spitzemapko, file = 'spitzemapko.rda')
```

Show that all arrays are deletions (SGA from dma, or from gsp1 emap, which has only deletions)
```{r}
spitzemapko %>%
  filter(!grepl('dma', array_strain_id),
         !grepl('emap', array_strain_id))
```

Find the intersecting array ORFs between the SGA and Gsp1 EMAP
```{r}
# find intersecting arrays between the Gsp1 EMAP and the SGA
gsp1_arrays <- 
  spitzemapko %>% 
  filter(interaction_network == 'gsp1_pEMAP') %>% 
  select(array_ORF) %>% 
  distinct() %>% 
  unlist()

sga_arrays <-
  spitzemapko %>% 
  filter(interaction_network != 'gsp1_pEMAP') %>% 
  select(array_ORF) %>% 
  distinct() %>% 
  unlist()

shared_arrays <- intersect(gsp1_arrays, sga_arrays)

print(length(gsp1_arrays))
print(length(sga_arrays))
print(length(shared_arrays))

# [1] 1444
# [1] 3827
# [1] 1137
```

We must use unique id's whenever spreading, otherwise there
will be multiple values for the same row-column pair in the
wide format matrix. The following code shows that when using
the strain_id's as unique id's, the number of distinct rows
is equal to the number of rows of spitzemap_for_corr.
```{r}
spitzemapko %>% 
  select(query_strain_id, array_ORF, score) %>%
  distinct() %>% 
  nrow() %>% 
  print()
print(nrow(spitzemapko))
# [1] 14375673
# [1] 14375673
```

The following full spreading (long => wide) works for the full matrix,
so the unique strain id's work. Note that this takes a while to run (5min)
```{r}
cormat <-
  spitzemapko %>% 
  select(query_strain_id, array_ORF, score) %>%
  spread(query_strain_id, score) %>% 
  select(-array_ORF) %>% 
  cor(., use = "pairwise.complete.obs", method = 'pearson')
```



