---
title: "gi_confidence_thresholds.Rmd"
author: "Chris Mathy"
email: "chris.mathy@ucsf.edu"
date: "3/22/2019"
output: html_document
---

```{r, setup, include=FALSE}
require('knitr')
opts_knit$set(root.dir = '../..')
```


import packages and data
```{r echo=T, results='hide'}
library(tidyverse)
load('Data/spitzemap.rda')
```

One sensible way to identify significance thresholds for EMAP data is to consider
Fig 4c from the following reference:

Collins, S. R., Schuldiner, M., Krogan, N. J., & Weissman, J. S. (2006). A strategy
for extracting and analyzing large-scale quantitative epistatic interaction data.
Genome biology, 7(7), R63.

This figure plots the estimated fraction of observations that arise from genuine
genetic interactions for a given averaged S-score (so the average score from both
kan/nat pair and nat/kan pair measurements. This fractional expectation gives a
measure of confidence that a given S-score reflects a genuine interaction, and
could be used as weights when performing analyses with S-scores.

Below, some points are estimated from figure 4c and used to fit a spline that
defines our confidence function. One can crudely overlay the transparent figure
plotted below onto the paper figure to show that we get fairly reasonable
agreement.

The fit spline produces the following values:
  -6    -5    -4    -3    -2    -1    0     1     2     3     4     5     6
  1.00  0.99  0.95  0.83  0.43  0.14  0.02  0.07  0.26  0.65  0.86  0.95  0.99
  
Note that the spline fits S-score < -5.6 to values slightly (by 1e-3) above 1.
If these confidence values are being used as weights for scores, simply cap all
weights at 1.0.
```{r}
# Estimate points from Fig 4c
d <- data.frame(avg_S_score = seq(-6, 6, by=0.5),
                confidence = c(1, 1, 0.99, 0.98, 0.95, 0.9, 0.85, 0.65, 0.45,
                               0.2, 0.15, 0.08, 0, 0.05, 0.07, 0.15,
                               0.22, 0.5, 0.65, 0.75, 0.87, 0.93, .95, .98, 0.99))

confidence_spline <- smooth.spline(x = d$avg_S_score, y = d$confidence, spar = 0.3)

confidence_fxn <- data.frame(x <- seq(-6, 6, by = 0.1),
                             y <- predict(confidence_spline, seq(-6, 6, by = 0.1))$y)


# S-score < -4 or > 4.9 has a confidence value greater than 0.95, using our spline
lower_sig_thresh <- max(confidence_fxn$x[(confidence_fxn$y > 0.95) &
                                         (confidence_fxn$x < 0)])
upper_sig_thresh <- min(confidence_fxn$x[(confidence_fxn$y > 0.95) &
                                         (confidence_fxn$x > 0)])
print(lower_sig_thresh)
print(upper_sig_thresh)


ggplot(confidence_fxn, aes(x=x, y=y)) +
  geom_point(color='red', size = 0.5) +
  geom_hline(yintercept = 0.95, color = 'blue') +
  xlab('Averaged S-score') + ylab('Confidence') + 
  theme_classic() +
  theme(
    text = element_text(family = "Helvetica", size = 6),
    axis.title = element_text(size = 6),
    axis.text = element_text(size = 6),
    axis.ticks = element_line(size = 0.05),
    axis.ticks.length = unit(0.05, 'cm'),
    axis.line = element_line(size = 0.1),
    panel.grid.major = element_line(colour = "grey50")
  )
ggsave('Supplemental_Figures/S_score_confidence_spline.png', height = 3, width = 3)
```

Additionally, The SGA dataset from Costanzo et al 2016 was published with
suggested thresholds:
  - lenient (P < 0.05)
  - intermediate (P < 0.05 and |score| > 0.08)
  - stringent (P <0.05 and score > 0.16 or score < -0.12)
  
For the intermediate confidence, the code below shows that the scaled score
thresholds corresponding to |score| > 0.08 are scaled_score < 2.25 and
scaled_score < -2.97.

```{r}
# SGA p-value filtering
spline_input <- read_delim('Supplemental_Figures/SGA_Scaling/scaler_output/spline.txt', delim = '\t')
scaling_spline <- smooth.spline(x = spline_input$score, y = spline_input$scale, spar = 0.02)

# compute scaled thresholds for intermediate stringency of SGA (abs(score) > 0.08)
adj = .0051 # need to add adjustment to score first (adjustment = .0051)
print(predict(scaling_spline, 0.08 + adj)$y * (0.08 + adj)) # upper threshold
print(predict(scaling_spline, -0.08 + adj)$y * (-0.08 + adj)) # lower threshold
```


One other question we might ask is: what is the distribution of pvalues for SGA
scores that have been scaled to EMAP format? The plot below bins values in
intervals of 1, so the boxplot for rounded_score of -1 corresponds to all scores
between -1.00 and -1.99. We see that the median p-value is below
0.05 (-log(p) > 1.3) for scores > 3 or < -5.
```{r}
spitzemap %>% 
  filter(interaction_network != 'gsp1_pEMAP', abs(score) < 10) %>%
  select(score, pvalue) %>%
  mutate('neg_log_pvalue' = -log10(pvalue),
         'score_rounded' = as.factor(trunc(score))) %>%
  ggplot(aes(x = score_rounded, y = neg_log_pvalue)) +
    geom_boxplot(outlier.shape = NA) +
    geom_hline(yintercept = -log10(0.05), color = 'red') +
    ylim(0, 2) +
    xlab('SGA score scaled to E-MAP format') + ylab('SGA p-value, from CellMap\n(negative log transformed)') +
    ggtitle('Distribution of SGA p-values scaled to EMAP format\nred line indicates a p-value of 0.05') +
    theme_classic() +
    theme(
      text = element_text(family = "Helvetica", size = 6),
      axis.title = element_text(size = 6),
      axis.text = element_text(size = 6),
      axis.ticks = element_line(size = 0.05),
      axis.ticks.length = unit(0.05, 'cm'),
      axis.line = element_line(size = 0.1),
      panel.grid.major = element_line(colour = "grey50")
    )
ggsave('Supplemental_Figures/SGA_scaled_score_vs_pval.pdf', height = 3, width = 3.5)
dev.off()
```















