---
title: "Scaling SGA2016 to EMAP Format"
author: "Chris Mathy"
email: "chris.mathy@ucsf.edu"
date: "2/28/2019"
output: html_document
last update: "3/5/2019"
---

```{r, setup, include=FALSE}
require('knitr')
opts_knit$set(root.dir = '../..')
```

Import packages -- this notebook makes use of the tidyverse functions,
version tidyverse_1.2.1
```{r}
library(tidyverse)
library(R.matlab)
```

Download SGA data (Costanzo 2016) from The Cell Map website (uncomment to run)
```{r}
# url <- 'http://boonelab.ccbr.utoronto.ca/supplement/costanzo2016/data_files/Data%20File%20S1_Raw%20genetic%20interaction%20datasets:%20Pair-wise%20interaction%20format.zip'
# download_dir <- dir.create('Data/SGA_Scaling/costanzo2016_raw_data/',showWarnings = FALSE)
# download.file(url=url, destfile='Data/SGA_Scaling/costanzo2016_raw_data/costanzo2016.zip')
# unzip('Data/SGA_Scaling/costanzo2016_raw_data/costanzo2016.zip', exdir='costanzo2016_data')
# file.rename(from='Data/SGA_Scaling/costanzo2016_raw_data/Data File S1. Raw genetic interaction datasets: Pair-wise interaction format',
#             to='costanzo2016_raw_data/datafileS1')
```

Read in each of the interaction networks included in the SGA, and make the
following ease-of-use data preprocessing steps:
- Column-wise concatenate the four tables into one table, and introduce a new
  column "interaction_network" identifying from which network each genetic
  interaction measurement originated.
- From the Strain ID columns, create "ORF" and "Descriptor" columns, and from
  the Allele Name columns, create "name" and "mutant" columns. Do this for
  both Query and Array columns. This is done by separating on "_" for strain
  IDs and on "-" for allele names. For example, "query_strain_id:
  YAL001C_tsq508" yields "query_ORF: YAL001C" and "query_descriptor: tsq508".
  Note that not many allele names, such as deletions, do no have mutant values
  (i.e. they don't contain "-"), and so NA's may be introduced in the
  "query_mutant" and "array_mutant" columns.
- Rename column names with spaces into a common format, e.g. "`Query Strain ID`"
  becomes "query_strain_id"
- Save the preprocessed SGA. Rdata files are well compressed and fast loading,
  Tab-delimited text files are bulky but readable.
```{r}
SGA_ExE <- read_delim('Data/SGA_Scaling/costanzo2016_raw_data/datafileS1/SGA_ExE.txt', delim='\t') # essential vs. essential
SGA_ExN <- read_delim('Data/SGA_Scaling/costanzo2016_raw_data/datafileS1/SGA_ExN.txt', delim='\t') # essential vs. non-essential
SGA_NxN <- read_delim('Data/SGA_Scaling/costanzo2016_raw_data/datafileS1/SGA_NxN.txt', delim='\t') # non-essential vs. non-essential
SGA_DAmP <- read_delim('Data/SGA_Scaling/costanzo2016_raw_data/datafileS1/SGA_DAmP.txt', delim='\t') # non-essential vs. non-essential

SGA <-
  bind_rows('ExE' = SGA_ExE,
            'ExN' = SGA_ExN,
            'NxN' = SGA_NxN,
            'DAmP' = SGA_DAmP,
            .id = 'interaction_network') %>%
  rename('query_strain_id' = `Query Strain ID`,
         'array_strain_id' = `Array Strain ID`,
         'query_allele_name' = `Query allele name`,
         'array_allele_name' = `Array allele name`,
         'arraytype_temp' = `Arraytype/Temp`,
         'score' = `Genetic interaction score (ε)`,
         'pvalue' = `P-value`,
         'query_smf' = `Query single mutant fitness (SMF)`,
         'array_smf' = `Array SMF`,
         'dmf' = `Double mutant fitness`,
         'dmf_sd' = `Double mutant fitness standard deviation`) %>% 
  separate(query_strain_id,into = c('query_ORF', 'query_descriptor'),
           sep ='_', extra = 'merge', remove = FALSE) %>%
  separate(array_strain_id, into = c('array_ORF', 'array_descriptor'),
           sep ='_', extra = 'merge', remove = FALSE) %>%
  separate(query_allele_name, into = c('query_name', 'query_mutant'),
           sep ='-', extra = 'merge', remove = FALSE) %>%
  separate(array_allele_name, into = c('array_name', 'array_mutant'),
           sep ='-', extra = 'merge', remove = FALSE) 

save(SGA, file='Data/SGA_Scaling/SGA_2016_full.rda') # Rdata
# write_delim(SGA, 'Data/SGA_Scaling/SGA_2016_full.txt', delim = '\t') # Tab-delimited text
```

Remove unprocessed data, leaving on the processed SGA (or if data is already
saved, load in the datafiles)
```{r}
# rm(list=setdiff(ls(), c('SGA')))
load('Data/SGA_Scaling/SGA_2016_full.rda')
```

Subsetting the SGA for analysis in conjunction with the Gsp1 pEMAP
- In addition to the NxN network, we include the ExN and ExE networks, as
  Costanzo et al 2016 reported having performed quantile normalization scaling
  to ensure that the networks were comparable. In short, mutants from the
  nonessential deletion mutant array (DMA) were included on the essential
  temperature sensitive arrays (TSA), providing ~297,000 measurements from
  mutants common to both arrays, and the mathematical transformation that
  correctly normalized the common mutants was applied to all TSA-derived scores
  (which populate the ExN and ExE networks).
- We discard the DAmP network because Costanzo et al 2016 reported significant
  sparseness of genetic interaction profiles of DAmP query strains compared to
  those of temperature sensitive alleles.
- Interactions involving an allele with the "supp" designaton are removed, as
  they denote strains that carry a second, spontaneous suppressor mutation.
- After these changes, the SGA kept 15,778,630 interactions out of an initial
  17,148,734 interactions, corresponding to a loss of only 8%.

Defining a unique ID for each strain, independent of marker:
- Both deletion and ts mutants can appear as either natMX (query) or kanMX
  (array) strains, and they are labeled differently
- For example, YLR166C_tsq66 and YLR166C_tsa63 are both the sec10-2 mutants,
  but their different strain IDs reflect their query/array status, determined
  by their marker.
- Below is a nice example, where YLR166C appears as both a ts and a DAmP
  perturbation (each natMX marked) in conjunction with a YAL002W array strain
  (kanMX marked), but then also as an array ts strain (kanMX marked) in
  conjuction with a YAL002W query strain (natMX marked). All five strains have
  unique strain IDs.
- Standard EMAP analysis averages measurements of gene pairs in which the
  perturbations are the same but the markers are swapped. To prepare the data
  for averaging by the EMAP Toolbox in MATLAB, we must identically label query
  and array strains that differ only by marker. Since we have seen that the
  strain_id values are marker dependent, we will give a new "unique_id" formed
  by combining the ORF and allele_name.
  
```{r}
# Example showing different strain IDs for the same mutant as query and array gene
SGA %>%
  filter((grepl('YLR166C', query_strain_id) & 
          grepl('YAL002W', array_strain_id)) |
         (grepl('YAL002W', query_strain_id) & 
          grepl('YLR166C', array_strain_id))) %>% 
  select(query_strain_id, query_allele_name, array_strain_id,
         array_allele_name, score)
```

Subsetting the SGA for downstream analysis (after scaling)
```{r}
SGA_unscaled <- SGA %>%
  filter(interaction_network %in% c('NxN', 'ExN', 'ExE')) %>% 
  filter(!grepl('supp', query_allele_name), !grepl('supp', array_allele_name)) %>% 
  mutate(query_unique_id = paste(query_ORF, query_allele_name, sep = '_'),
         array_unique_id = paste(array_ORF, array_allele_name, sep = '_'))

write_delim(SGA_unscaled, 'Data/SGA_Scaling/SGA_unscaled_long.txt', delim = '\t')
```

Confirmation that the unique IDs result in no duplicate interaction measurements
```{r}
SGA_unscaled %>%
  select(query_unique_id, array_unique_id, score) %>% 
  group_by(query_unique_id, array_unique_id) %>% 
  mutate(n_rep = n()) %>% 
  ungroup() %>% 
  group_by(n_rep) %>%
  summarise(n())
```

Confirmation that there are still gene pairs with duplicate measurements for
swapped markers
```{r}
SGA_unscaled %>%
  filter(query_unique_id %in% c('YLR166C_sec10-2','YAL002W_vps8')) %>% 
  filter(array_unique_id %in% c('YLR166C_sec10-2','YAL002W_vps8')) %>% 
  select(query_unique_id, array_unique_id, score)
```


Subsetting the SGA for scaling to a reference EMAP:
- To improve the accuracy of comparison between the SGA and the Gsp1 pEMAP, the
  SGA will be scaled to EMAP format according to the method described in
  Ryan et al 2012 Molecular Cell.
- The chromosome biology EMAP ('cF3', Collins et al 2007 Nature) will serve as
  the reference EMAP, as in Ryan et al 2012.
- The method seeks to fits a nonlinear scaling function so that interactions
  common to two genetic interaction datasets are scored similarly.
- Most genes in the cF3 EMAP are deletions or DAmPs, though there are a few
  pseudogene strains and temperature sensitive mutants. Non-deletions are
  not be included in the scaling. As such, only the NxN SGA is needed for
  scaling, as non of the ExN or ExE gene pairs are present in the cF3 EMAP

Performing the scaling
- The NxN network as well as the full SGA are converted from long to 
  wide format and undergo averaging in the MATLAB EMAP Toolbox, which
  averages replicates of gene pairs with swapped markers, and performs
  a pseudo-averaging for singlets. This averaging step was done for the
  Gsp1 pEMAP and the cF3 EMAP.
- The NxN is scaled to the cF3 EMAP, resulting in a non-linear scaling spline.
- The scaling spline is applied to the full, averaged SGA, resulting in a
  final dataset with scores more comparable to the Gsp1 pEMAP.

Spread the NxN network into wide format for MATLAB averaging/pseudo-averaging
```{r}
# For the NxN network, the ORFs can be used to spread, as they function
# as unique identifiers.
SGA_NxN_spread <-
  SGA_unscaled %>% 
  filter(interaction_network == 'NxN') %>% 
  select(query_ORF, array_ORF, score) %>% 
  spread(array_ORF, score)

# create the necessary objects for the pseudo-averaging script in the EMAP MATLAB toolbox

SGA_NxN_unaveraged <- list()
SGA_NxN_unaveraged$rowlabels <- SGA_NxN_spread$query_ORF
SGA_NxN_unaveraged$collabels <- colnames(SGA_NxN_spread)[-1]
SGA_NxN_unaveraged$data <- select(SGA_NxN_spread, -query_ORF)
colnames(SGA_NxN_unaveraged$data) <- NULL

writeMat(con="Data/SGA_Scaling/SGA_NxN_unaveraged.mat", SGA_NxN_unaveraged = SGA_NxN_unaveraged)

```


Spread the full SGA into wide format for MATLAB averaging/pseudo-averaging
```{r}
SGA_spread <-
  SGA_unscaled %>% 
  select(query_unique_id, array_unique_id, score) %>% 
  spread(array_unique_id, score)

SGA_unaveraged <- list()
SGA_unaveraged$rowlabels <- SGA_spread$query_unique_id
SGA_unaveraged$collabels <- colnames(SGA_spread)[-1]
SGA_unaveraged$data <- select(SGA_spread, -query_unique_id)
colnames(SGA_unaveraged$data) <- NULL

writeMat(con="Data/SGA_Scaling/SGA_unaveraged.mat", SGA_unaveraged = SGA_unaveraged)
```


At this step, the matlab files are imported into MATLAB and run through the
averaging/pseudo-averaging step. The output matrix of the SGA_NxN subset is
then fed into the scaling script (written in python) and a scaling spline
that scales the SGA to the cF3 EMAP is calculated. This curve is saved
as a series of points, and the same spline can be fit for use in R.

```{r}



spline_input <- read_delim('Supplemental_Figures/SGA_Scaling/scaler_output/spline.txt',delim = '\t')
scaling_spline <- smooth.spline(x = spline_input$score,
                                y = spline_input$scale,
                                spar = 0.02)
plot(scaling_spline)
```


Get adjustment of unscaled values from the scaler logfile. The adjustment was
applied before scaling so that there were the same number of negatives in each
dataset. As such, the scaling spline is only accurate when used after this
slight adjustment.

Also get the min and max scores from the dataset used for scaling. As done in the
Ryan et al 2012 scaling, any scores more negative than the min score or more
positive than the max score used for scaling use the scaling factors of the min or
max. In practice, if a value to be scaled is larger than any value in our training
set, we use the scaling factor from the largest observed value of the training set.
```{r}
adjustment <-
  grep('adjustment=', readLines('Supplemental_Figures/SGA_Scaling/scaler_output/scaler_log.txt'), value=TRUE) %>% 
  strsplit(., '=') %>% 
  `[[`(1) %>% 
  `[`(2) %>% 
  as.numeric()

max_score <-
  grep('max_score=', readLines('Supplemental_Figures/SGA_Scaling/scaler_output/scaler_log.txt'), value=TRUE) %>% 
  strsplit(., '=') %>% 
  `[[`(1) %>% 
  `[`(2) %>% 
  as.numeric()

min_score <-
  grep('min_score=', readLines('Supplemental_Figures/SGA_Scaling/scaler_output/scaler_log.txt'), value=TRUE) %>% 
  strsplit(., '=') %>% 
  `[[`(1) %>% 
  `[`(2) %>% 
  as.numeric()

# upper/lower bounds scaling factors
ubsf <- predict(scaling_spline, max_score)$y
lbsf <- predict(scaling_spline, min_score)$y
```

The scaling is performed according to the method employed by Ryan et al 2012,
shown in scaler.py and below (code received through personal communication).

#-----from scaler.py-----#
# This function creates a curve which maps scores to scaling factors
# The s=0.02 defines how close the curve fits your data points
s = UnivariateSpline(score, scale, s=0.02)

# if the value to be scaled is larger than any value in our training set,
# we use the scaling factor from the largest observed value
def s_bounded(x) :
    if x<min(score) :
        x = min(score)
    elif x > max(score) :
        x=max(score)
    return s(x)
    
# This function applies our scaling factor to a given value
g = lambda x : (x + adjustment) * s_bounded(x + adjustment)


Scale the SGA
```{r}
SGA_scaled_to_EMAP <- SGA %>%
  mutate(score_adjusted = score + adjustment) %>% 
  mutate(scaling_factor = case_when(score_adjusted > max_score ~ ubsf,
                                    score_adjusted < min_score ~ lbsf,
                                    TRUE ~ predict(scaling_spline, .$score_adjusted)$y)) %>% 
  mutate(score_scaled_to_emap = score_adjusted * scaling_factor)
```

Save the scaled_SGA
```{r}
save(SGA_scaled_to_EMAP, file='Data/SGA_Scaling/SGA_2016_full_scaled.rda')
```

Load the scaled_SGA
```{r}
load('Data/SGA_Scaling/SGA_2016_full_scaled.rda')
```

Example of how to get all interactions involving a gene, in this case Gsp1
```{r}
SGA_scaled_to_EMAP %>%
  filter(grepl('gsp1', query_allele_name) | grepl('gsp1', array_allele_name)) %>% 
  select(query_strain_id, query_allele_name,
         array_strain_id, array_allele_name, score_scaled_to_emap) %>% 
  arrange(score_scaled_to_emap)

```







